---
title: "Creating EML Using EDIutils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating EML Using EDIutils}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
The following libraries are needed to create a working EML document. 
```{r, message = FALSE}
library(EDIutils)
library(tidyverse)
library(readxl)
library(EML)
```
In this document, we can quickly create an EML file which has the opportunity to 
append the following elements. For more detailed instructions on a specific section 
please see section specific articles. 

```yaml
- eml 
  -access
  -dataset
    -personnel 
      - creator
      - contact
      - associated party 
    - title  
    - abstract 
    - keyword set  
    - license 
    - methods 
    - maintenance 
    - project
      - project title
      - project personnel 
      - project funding 
    - coverage 
      - geographic coverage 
      - temporal coverage 
      - taxonomic coverage 
    - data table 
      - physical 
      - attribute list 
      - methods 
```

## Our Example 
We will use the data provided by John Hannon to show an example of how to put the
metadata together in a way which can be uploaded to the EDI data portal website. 
The excel sheet of which Hannon's data resides is found in "~/data-raw/Hannon-Example/example-metadata.xlsx". 

## Required Files 
Four files must be uploaded to make a complete EML document. 

* An excel spreadsheet containing the majority of the metadata
* A word document containing the abstract text 
* A word document containing the methods text 
* A file that contains the data (preferably csv if tabular data) and the name of the file 

In addition to these four files a entity description and an EDI number must be defined. 
The EDI number should be a unique data package identifier. You can reserve this 
data package identifier number on the [EDI data repository under tools.](https://portal.edirepository.org/nis/reservations.jsp)
For this example all of our files are stored in a the Hannon-Example folder within data-raw. 
"~/data-raw/Hannon-Example".

```{r, include = FALSE}
excel_path <- system.file("extdata", "Hannon-Example", "example-metadata.xlsx", 
                          package = "EDIutils", mustWork = TRUE)

sheets <- readxl::excel_sheets(excel_path)
metadata <- lapply(sheets, function(x) readxl::read_excel(excel_path, sheet = x))
names(metadata) <- sheets

abstract_docx <- system.file("extdata", "Hannon-Example", "hannon_example_abstract.docx", 
                             package = "EDIutils", mustWork = TRUE)

methods_docx <- system.file("extdata", "Hannon-Example", "hannon_example_methods.docx", 
                            package = "EDIutils", mustWork = TRUE)

dataset_file <- system.file("extdata", "Hannon-Example", "hannon_example_physical_data.csv",
                               package = "EDIutils", mustWork = TRUE)

dataset_file_name <- "hannon_example_physical_data.csv"

edi_number <- "edi.678.1"
```

```{r, eval = FALSE}
excel_path <- "data-raw/Hannon-Example/example-metadata.xlsx"

sheets <- readxl::excel_sheets(excel_path)
metadata <- lapply(sheets, function(x) readxl::read_excel(excel_path, sheet = x))
names(metadata) <- sheets

abstract_docx <- "data-raw/Hannon-Example/hannon_example_abstract.docx"

methods_docx <- "data-raw/Hannon-Example/hannon_example_methods.docx"

dataset_file <- "data-raw/Hannon-Example/hannon_example_physical_data.csv"

dataset_file_name <- "hannon_example_physical_data.csv"

edi_number <- "edi.678.1"
```
## EML Sections 
We start with the simpler pieces of information such as `personnel` and `title` and 
work up to appending the more complex sections like `dataTable`. 

### Access Permissions 
The `add_access` adds an access section at the beginning of our EML document. The 
`add_access` default is public principal with a read permission. 

```{r}
access <- add_access()
```

### Publication Date 
The publication date does not have a tab in the excel sheet, but is simply added with 
the `add_pub_date` function. If no date is provided, it will automatically append
the current date. This can be overwritten by providing an input for `date`. 

```{r}
pub_date <- add_pub_date(list())
```

### Personnel: Creator, Contact, and Associated Parties. 
The `add_personnel` function allows you to append information on the creator or
the associated parties to your file. The creator would be any person or organization
who is responsible for the creation of the data, while an associated party would 
be any person or organization which is associated to the dataset. The creator will also 
be the contact for this datset. Multiple creators, associated parties, and a mixture
of both can be appended to the EML file by providing all inputs in the "personnel"
tab in the "example-metadata.xlsx" excel file. 

```{r}
personnel <- list()
adds_person <- function(first_name, last_name, email, role, organization, orcid) {
  personnel <- add_personnel(personnel, first_name, last_name, email,
                                  role, organization, orcid)
}
personnel <- purrr::pmap(metadata$personnel, adds_person) %>% flatten()
```
### Title and Short Name 
Following the personnel element, we can append the title elements. The `add_title` function 
allows you to append the title and short name of the dataset to the file. 
Only one title and short name should be appended. The title should be 
fairly descriptive and between 7 and 20 words long. The short name must be less than
the number of words present in the title and is your opportunity to give viewers a more 
accessible name to the dataset. Again, this information can be added to the "title"
tab in the excel file "example-metadata.xlsx".

```{r, }
title <- add_title(list(), title = metadata$title$title,  
                   short_name = metadata$title$short_name)
```

### Keyword Set 
The next item we will append is the keyword set. We can use the `add_keyword_set` 
function to do so. The keyword set should include a list of words that help 
identify your project or connect it with other similar projects. If you add keywords
from a controlled vocabulary please see the detailed instructions on 
[initial metadata information] (vignettes/Initial-Metadata-Information).

```{r}
keywords <- add_keyword_set(list(), metadata$keyword_set[,1])
```

### Abstract
Next, we will use the `add_abstract` function to append the abstract of the dataset to 
your file. The abstract should include basic information on the dataset that gives 
a brief summary to the viewers of what they are observing from the data. This 
information will not be added to the "example-metadata.xlsx" excel sheet, but rather the 
"hannon_example_abstract.docx". The text provided can be deleted and replaced with your own. 

```{r}
abstract <- add_abstract(list(), abstract = abstract_docx)
```

### License and Intellectual Rights
Following the abstract, we will append the license and intellectual rights information. 
The `add_license` function allows you to append the licensing and usage information 
to your file. If you are not adding a CC0 or CCBY default license please see [initial metadata information] (vignettes/Initial-Metadata-Information). This information can be added to the 
"license" tab in the excel file "example-metadata.xlsx".

```{r}
license <- add_license(list(), default_license = metadata$license$default_license)
```

### Methods
The method section explains the scientific methods that were used in the collection of the dataset.
The `add_method` function allows you to add a method file to the parent_element.
A template methods document is given ("data-raw/Hannon-Example/hannon_example_methods.docx"). 
If more methods are needed, you can create separate sections in your word document.

```{r}
methods <- add_method(list(), methods_file = methods_docx)
```

### Maintenance 
The maintenance of a dataset \strong(is simply if the data collection is complete or ongoing.)
The `add_maintenance` function allows you to append the status of the dataset to
your file and the inputs of `complete` and `ongoing` are the only ones allowed.
If the dataset is still in progress, the frequency of which it is updated must 
be provided as well. This information should be added to the "maintenance" tab 
in the excel file "example-metadata.xlsx".

```{r}
maintenance <- add_maintenance(list(), status = metadata$maintenance$status,
                               update_frequency = metadata$maintenance$update_frequency)
```

### Project: Title, Personnel, and Funding 
The project section should be appended next. Project personnel and project funding 
are nested in this project section. To generate a valid EML doc the project section 
must contain a project title, a project personnel, and project funding. 

#### Project title
```{r}
project_title <- metadata$project$project_title
```


#### Project personnel 
```{r}
project_table <- metadata$project %>% select(-project_title)
project_personnel <- list()
adds_project_person <- function(first_name, last_name, email, role, organization, orcid) {
  project_personnel <- add_personnel(project_personnel, first_name, last_name, email,
                                  role, organization, orcid)
  if (!is.null(project_personnel$associatedParty)) {
  project_personnel <- project_personnel$associatedParty
  }
  if (!is.null(project_personnel$creator)){
  project_personnel <- project_personnel$creator
  }
  return(project_personnel)
}
project_personnel <- purrr::pmap(project_table, adds_project_person) %>% flatten()
```

#### Project funding 
The `add_funding` function allows you to append 
both the description of the funding you have received as well as the organization 
you received the funding from. Multiple funders can be appended if necessary and 
information should be added to the "funding" tab in the excel file "example-metadata.xlsx".

```{r}
add_a_funder <- function(funder_name, funder_identifier, award_number, award_title, 
                         award_url, funding_description) {
  award_information <- add_funding(funder_name, funder_identifier, award_number, award_title,
                                award_url, funding_description)
}

award_information <- purrr::pmap(metadata$funding, add_a_funder) %>% flatten()
```

#### Combining Project Elements
Once all the components of project have been defined we can use the add_project
function to combine the sections in the proper formatting for an EML document. 
```{r}
project <- add_project(list(),
                       project_title, 
                       award_information,
                       project_personnel)
``` 

### Coverage: Geographic, Temporal, Taxonomic 
Next, the coverage information is appended. The `add_coverage` function allows
you to append full coverage information to your file. Temporal, Geographic, and 
Taxonomic Coverage are all required elements to generate an EML document. 

##### Taxonomic coverage 
Taxonomic coverage can be appended using the `add_taxonomic_coverage` function. 
The taxonomic coverage information is added to the "example-metadata.xlsx" 
excel file on the tab "taxonomic_coverage". chinook, delta_smelt, steelhead, 
white_sturgeon, and green_sturgeon are all default options that can be selected 
using the drop down menu under the CVPIA_common_species column.
```{r}
taxonomic_coverage <- purrr::pmap(metadata$taxonomic_coverage, add_taxonomic_coverage)
```

#### Combining Coverage Elements  
The `add_coverage` function will add all the elements of coverage to the parent 
element. 
```{r}
coverage <- add_coverage(list(),
                         geographic_description = metadata$coverage$geographic_description,
                         west_bounding_coordinate = metadata$coverage$west_bounding_coordinate,
                         east_bounding_coordinate = metadata$coverage$east_bounding_coordinate,
                         north_bounding_coordinate = metadata$coverage$north_bounding_coordinate,
                         south_bounding_coordinate = metadata$coverage$south_bounding_coordinate,
                         begin_date = metadata$coverage$begin_date,
                         end_date = metadata$coverage$end_date,
                         taxonomic_coverage = taxonomic_coverage)

```

### Data Table 
Next, we need to create the data table. The data table element includes`physical`, `attribute_list`, 
and potentially additional information for Spatial Data. These sections are all lists 
which must be created first, appended to a dataTable, and then appended to 
the `parent_element`.

#### Physical 
The physical element can be created first using the `add_physical` function. 
This will append information on the actual physical information of the data. 
The file path of the data can be added to the "example-metadata.xlsx" excel file 
and all parameters will be appended, gathering the csv's unique size and 
authentication, as well as providing the default physical values of a csv. 

```{r}
physical <- add_physical(file_path = dataset_file)
```

If you wish to provide alternative information than the defaults of a csv, you must do
so manually in Rstudio using the `add_physical` function fully. It is not recommended 
that you do so, however, as the default values present represent the best practices 
in most cases. 

#### Attribute List 
Next, we will need to append an attribute list to the data table. We will be able 
to use the `add_attribute` function to do so. Please make sure you review what 
type of attribute you are providing and what inputs are necessary. These values 
can then be inputted into the "attribute" tab in the "example-metadata.xlsx" excel file. If you
are using a "nominal" or "ordinal" attribute which is "enumerated", (it has a specific 
code definition), please also use the tab "code_definitions". Provide each
unique code and definition, with its `attribute_name` aligning to that which is in the
"attribute" tab. An example is present currently to help better showcase this. Every single 
column in the dataTable must have a described attribute to match EDI congruence checker. 

```{r}
attribute_table <- metadata$attribute
codes <- metadata$code_definitions
attribute_list <- list()

code_helper <- function(code, definitions) {
  codeDefinition <- list(code = code, definition = definitions)
}

attributes_and_codes <- function(attribute_name, attribute_definition, storage_type, 
                                 measurement_scale, domain, type, units, unit_precision, 
                                 number_type, date_time_format, date_time_precision, minimum, maximum, attribute_label){
  if (domain %in% "enumerated") { 
    definition <- list()
    current_codes <- codes[codes$attribute_name == attribute_name, ]
    definition$codeDefinition <- purrr::pmap(current_codes %>% select(-attribute_name), code_helper) 
  } else {
    definition = attribute_definition
  }
  new_attribute <- add_attribute(attribute_name = attribute_name, attribute_definition = attribute_definition,
                                 storage_type = storage_type, measurement_scale = measurement_scale, domain = domain,
                                 definition = definition, type = type, units = units, unit_precision = unit_precision,
                                 number_type = number_type, date_time_format = date_time_format,
                                 date_time_precision = date_time_precision, minimum = minimum, maximum = maximum,
                                 attribute_label = attribute_label)
}

attribute_list$attribute <- purrr::pmap(attribute_table, attributes_and_codes) 
```
#### Putting the Data Table Together
Now that we have the `attribute_list` and `physical` information, we 
can compose the data table, which is the last element we need to create our 
working EML file. If you wish to append multiple datatables to one datapackage please 
contact CVPIA data guidance and we will help you do so properly. 

```{r}
dataTable = list(entityName = dataset_file_name,
                 entityDescription = metadata$dataset$name,
                 physical = physical,
                 attributeList = attribute_list)

```
## Append all items to the dataset
```{r}
dataset <- list(title = title$title,
                shortName = title$shortName,
                creator = personnel$creator,
                contact = personnel$contact,
                pubDate = pub_date,
                abstract = abstract$abstract,
                associatedParty = list(personnel[[3]], personnel[[4]], personnel[[5]]),
                keywordSet = keywords$keywordSet,
                coverage = coverage$coverage,
                intellectualRights = license$intellectualRights,
                licensed = license$licensed,
                methods = methods,
                maintenance = maintenance$maintenance,
                dataTable = dataTable) 

```
## Making the EML document 
Now that we have all of the items appended to the `parent_element`, we must append it 
to the "dataset" and also give the file a unique ID number. The unique id number should 
match one that you reserved on the EDI data portal. Input this at the top of the 
document where you define the data tables. 
```{r}
eml <- list(packageId = edi_number,
            system = "EDI",
            access = access,
            dataset = dataset)
```
We now have all necessary elements to create a working EML document. The final step is to convert it into the correct format. To do so, we can use the EML library's `write_eml` function. 
```{r, eval = FALSE}
file_name <- paste(edi_number, "xml", sep = ".")
EML::write_eml(eml, file_name)
eml_validate(file_name)

```
