---
title: "Create EML from Template"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{create-EML-from-template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

This article demonstrates how to create an EML document for a data package containing multiple data entities. To follow along with this example, please download the [Banet-Example](https://cvpia-data-stewardship.s3-us-west-1.amazonaws.com/Banet-Example.zip). 

We will create a nested list from our metadata templates and then use the [EML](https://docs.ropensci.org/EML/) R package's [`write_eml`](https://docs.ropensci.org/EML/reference/write_eml.html) function to convert our list into a valid EML document. 

Our EML file will contain the following elements:

```yaml
- eml 
  -access
  -dataset
    - creator
    - contact
    - associated parties
    - title  
    - abstract 
    - keyword set  
    - license 
    - methods 
    - maintenance 
    - project
    - coverage 
    - data table 
```

We have two main sections within our EML document:

1. Access - The [Access Section](https://eml.ecoinformatics.org/schema/eml_xsd.html#eml_access) defines the access for the entire resource (entire EML document)
2. Dataset - The [Dataset Resource](https://eml.ecoinformatics.org/schema/eml_xsd.html#eml_dataset) contains the descriptive metadata information on a data set. A data set can contain one or more data entities. A data entity is a tabular data table or a vector or raster image. 

For a more in depth description of EML please see the [EML Specification.](https://eml.ecoinformatics.org/)


## Required R Packages
The following libraries are needed to create a working EML document. 
```{r}
library(EMLaide)
library(tidyverse)
library(EDIutils)
library(readxl)
library(EML)
```


## Datatable(s) and associated metadata

To follow along with this example please download the files from the [Banet-Example](https://cvpia-data-stewardship.s3-us-west-1.amazonaws.com/Banet-Example.zip).
Within the directory, there are two subdirectories ("data" and "metadata") which contain all the necessary data and metadata to create a valid EML document using EMLaide.

At minimum, four files are needed to use our tools:

1. **Dataset** text file containing machine readable tabular data 
2. **Metadata Template** An [excel workbook](https://cvpia-data-stewardship.s3-us-west-1.amazonaws.com/metadata+template.zip) containing the majority of the metadata. Full documentation on the template materials are [here](https://cvpia-osc.github.io/EMLaide/articles/template-doc.html).  
3. **Abstract Document** A word or markdown document containing the abstract text 
4. **Methods Document** A word or markdown document containing the methods text 

Because our data set is composed of multiple data entities, we are creating a dataframe with each row representing a different data entity with the following information:

  * *filepath* - filepath of the data entity (datatables)
  * *attribute_info* - filepath to the metadata describing the specific data entity (should be a metadata.xlsx)
  * *datatable_description* - brief description of data entity 
  * *datattable_url* - (Optional) url to a publicly accessible file hosted online. Required if using `EMLaide::evaluate_edi_package()` or `EMLaide::upload_edi_package()` to evaluate or upload your EML document to EDI from R. 
  * *datatable_methods* - (Optional) filepath data entity specific methods document
  * *additional_information* - (Optional) brief text description containing additional information that is pertinent to a specific data entity. 

This dataframe is the input to the [`add_datatable`](https://cvpia-osc.github.io/EMLaide/reference/add_datatable.html) function which function generates [attribute metadata](https://cvpia-osc.github.io/EMLaide/reference/create_attribute.html) from the *atribute_info* files and [physical information](https://cvpia-osc.github.io/EMLaide/reference/create_physical.html) describing the datatable from the *filepath* and *datatable_url* information. 

**Example Dataframe Structure for `datatable_metadata`** 

```{r, eval = TRUE}
datatable_metadata <- 
  dplyr::tibble(filepath = c("data/enclosure-study-growth-rate-data.csv",
                             "data/enclosure-study-gut-contents-data.csv",
                             "data/microhabitat-use-data-2018-2020.csv",
                             "data/seining-weight-lengths-2018-2020.csv",
                             "data/snorkel-index-data-2015-2020.csv"),  
                attribute_info = c("metadata/enclosure-study-growth-rates-metadata.xlsx",
                                   "metadata/enclosure-study-gut-contents-metadata.xlsx",
                                   "metadata/microhabitat-use-metadata.xlsx",
                                   "metadata/seining-weight-length-metadata.xlsx",
                                   "metadata/snorkel-index-metadata.xlsx"),
                datatable_description = c("Growth Rates - Enclosure Study",
                                          "Gut Contents - Enclosure Study",
                                          "Microhabitat Data",
                                          "Seining Weight Lengths Data",
                                          "Snorkel Survey Data"),
                datatable_url = paste0("https://raw.githubusercontent.com/FlowWest/edi.749.1/main/data/",
                                       c("enclosure-study-growth-rate-data.csv",
                                         "enclosure-study-gut-contents-data.csv",
                                         "microhabitat-use-data-2018-2020.csv",
                                         "seining-weight-lengths-2018-2020.csv",
                                         "snorkel-index-data-2015-2020.csv")))
```

Each row contains all information needed for a data entity to be added to the dataset element of a data package. If you only have one datatable keep this structure or use a named list with the same information.

```{r, eval=TRUE, echo=FALSE}

knitr::kable(datatable_metadata)
```

```{r, include = FALSE, eval= TRUE}
datatable_metadata <- 
  dplyr::tibble(filepath = c(system.file("extdata", "Banet-Example", "data", "enclosure-study-growth-rate-data.csv", 
                          package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "data", "enclosure-study-gut-contents-data.csv", 
                          package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "data", "microhabitat-use-data-2018-2020.csv", 
                          package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "data", "seining-weight-lengths-2018-2020.csv", 
                          package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "data", "snorkel-index-data-2015-2020.csv", 
                          package = "EMLaide", mustWork = TRUE)),  
                attribute_info = c(system.file("extdata", "Banet-Example", 
                                               "metadata", "enclosure-study-growth-rates-metadata.xlsx", 
                                               package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "metadata", "enclosure-study-gut-contents-metadata.xlsx", 
                          package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "metadata", "microhabitat-use-metadata.xlsx", 
                          package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "metadata", "seining-weight-length-metadata.xlsx", 
                          package = "EMLaide", mustWork = TRUE),
                          system.file("extdata", "Banet-Example", "metadata", "snorkel-index-metadata.xlsx", 
                          package = "EMLaide", mustWork = TRUE)),
                datatable_description = c("Growth Rates - Enclosure Study",
                                          "Gut Contents - Enclosure Study",
                                          "Microhabitat Data",
                                          "Seining Weight Lengths Data",
                                          "Snorkel Survey Data"),
                datatable_url = paste0("https://raw.githubusercontent.com/FlowWest/edi.749.1/main/data/",
                                       c("enclosure-study-growth-rate-data.csv",
                                         "enclosure-study-gut-contents-data.csv",
                                         "microhabitat-use-data-2018-2020.csv",
                                         "seining-weight-lengths-2018-2020.csv",
                                         "snorkel-index-data-2015-2020.csv")))

```



```{r, include = FALSE, eval = TRUE}
excel_path <- system.file("extdata", "Banet-Example", "metadata", "data-package-metadata.xlsx", 
                          package = "EMLaide", mustWork = TRUE) 
sheets <- readxl::excel_sheets(excel_path)
metadata <- purrr::map(sheets, function(x) readxl::read_excel(excel_path, sheet = x))
names(metadata) <- sheets

abstract_docx <- system.file("extdata", "Banet-Example", "metadata","abstract.docx", 
                             package = "EMLaide", mustWork = TRUE) 

methods_docx <- system.file("extdata", "Banet-Example", "metadata", "methods.docx", 
                            package = "EMLaide", mustWork = TRUE)  

```

### Data Package Metadata, Methods, and Abstract 

The following code loads the "data-package-metadata.xlsx", "abstrac.docx", and "methods.docx". Each sheet of the excel workbook pertains to a different metadata element and will be the input to the `add_[blank]` functions used throughout this example.  
```{r, eval = FALSE}
excel_path <- "Banet-Example/metadata/data-package-metadata.xlsx"
sheets <- readxl::excel_sheets(excel_path)
metadata <- lapply(sheets, function(x) readxl::read_excel(excel_path, sheet = x))
names(metadata) <- sheets

abstract_docx <- "metadata/abstract.docx"
methods_docx <- "metadata/methods.docx"
```


### EDI Identifier
In addition to these files, we will need a unique EDI data package identifier.  We use the function [`reserve_edi_id`](https://cvpia-osc.github.io/EMLaide/reference/reserve_edi_id.html) to generate a EDI id. You must already have an account associated with EDI to do this. 
```{r, eval = FALSE}
edi_number <- reserve_edi_id(user_id = "your user id", password = "your user password ")
```
You can also reserve this data package identifier on the [EDI data repository under tools](https://portal.edirepository.org/nis/reservations.jsp).

For this example, we will use the following identifier.
```{r}
edi_number <- "edi.750.1" 
```


## Create Dataset Element

We will use `magrittr::%>%` with our `add_[blank]` functions to append each EML element to a list. The `%>%` is a pipe like operator which takes the left-hand side as the first argument of the function appearing on the right-hand side.

For details on appropriate inputs to the functions see documentation at `?add_[blank]`.

The `add_methods()` and `add_abstract()` functions take in the `methods_docx` and the `abstract_docx`. The `add_datatable()` function takes in the `datatable_metadata` defined and described above. Every other function takes in one or more sheets from the `metadata` object. For template items with multiple rows, the `add_[blank]` functions map through each row and adds a named nested list for each row to the dataset element. 

The code below adds all dataset elements. 
```{r}
dataset <- list() %>%
  add_pub_date() %>%
  add_title(metadata$title) %>%
  add_personnel(metadata$personnel) %>%
  add_keyword_set(metadata$keyword_set) %>%
  add_abstract(abstract_docx) %>%
  add_license(metadata$license) %>%
  add_method(methods_docx) %>%
  add_maintenance(metadata$maintenance) %>%
  add_project(metadata$funding) %>%
  add_coverage(metadata$coverage, metadata$taxonomic_coverage) %>%
  add_datatable(datatable_metadata)
```

## Handle Custom Units

When units aren't standard `add_datatable()`will give a message like the following: `"We identified the following custom unit: fishPerSchool , please make sure to add information on this custom unit in additional metadata information:"`. We must formally define each of these custom units and add them to the EML document as an additional metadata section.

The code below defines 4 custom units and uses the `EML::set_unitList()` function to format them into a unitList that can be added to our EML document. 
```{r}
custom_units <- data.frame(id = c("fishPerEnclosure", "thermal unit", "day", "fishPerSchool"),
                           unitType = c("density", "temperature", "dimensionless", "density"),
                           parentSI = c(NA, NA, NA, NA),
                           multiplierToSI = c(NA, NA, NA, NA),
                           description = c("Fish density in the enclosure, number of fish in total enclosure space",
                                           "thermal unit of energy given off of fish",
                                           "count of number of days that go by",
                                           "Number of fish counted per school"))

unitList <- EML::set_unitList(custom_units)
```

## Combine EML Elements and Build Document 

The code below adds all of the elements we generated above and an access element into an `eml` list. 

* The `add_access` adds an access section at the beginning of our EML document. The 
`add_access` default is public principal with a read permission. 
* The `dataset` list from above contains all elements of the `dataset` section of the EML. This includes the `datatables`, `abstract`, `methods`, and all the other metadata sections appended above. 
* The `addtionalMetadata` contains the `unitList` that we generated to hold our custom units. 

```{r}
eml <- list(packageId = edi_number,
            system = "EDI",
            access = add_access(),
            dataset = dataset,
            additionalMetadata = list(metadata = list(unitList = unitList)))
```

Once all of our information is appended to our eml list we can use the [`write_eml`](https://docs.ropensci.org/EML/reference/write_eml.html) and [`eml_validate`](https://docs.ropensci.org/EML/reference/eml_validate.html) functions from the [EML](https://docs.ropensci.org/EML/index.html) package to convert our list to EML and check validity.
```{r}
EML::write_eml(eml, "edi.750.1.xml")
EML::eml_validate("edi.750.1.xml")
```

## Evaluation using the EDI's EML Congruency Checker
To evaluate your document in R using [EDI's EML Congruence Checker](https://github.com/EDIorg/ecc) you can use `evaluate_edi_package()`. To use this function you must have the data entities text files publically accessible by a URL. This URL must be added in the `datatable_metadata` section above. If you do not have a URL available then you can upload the EML document and the dataset on the [EDI data portal.](https://portal.edirepository.org/nis/home.jsp) 

```{r, eval = FALSE}
evaluate_edi_package(user_id = "Your User Id", password = "Your password", eml_file_path = "edi.750.1.xml")
```
