---
title: "Creating EML Using cvpiaEDIutils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating EML Using cvpiaEDIutils}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The following libraries are needed to create a working EML document. 
```{r, message = FALSE}
library(cvpiaEDIutils)
library(tidyverse)
library(readxl)
library(EML)
```
In this document, we create an EML file which has the opportunity to append the 
following elements:

```yaml
- eml 
  -dataset
    - creator
    - contact
    - associated party 
    - title 
    - short name 
    - abstract 
    - keyword set 
    - intellectual rights 
    - license 
    - methods 
    - maintenance 
    - coverage 
      - geographic coverage 
      - temporal coverage 
      - taxonomic coverage 
    - data table 
      - physical 
      - attribute list 
      - methods 
```

## Our Example 
We will use the data provided by John Hannon to show an example of how to put the
metadata together in a way which can be uploaded to the EDI data portal website. 
The excel sheet of which Hannon's data resides is found in "~/data-raw/template/template.xlsx". 
This file can be edited with your data and used with this document to create your 
own working EML file. 

## Creating EML 
The tabs of the excel file are organized in the order of which this template appends information. 
We start with the simpler pieces of information such as `personnel` and `title` and 
work up to appending the data table, which has other components to it. 

### Creating the Parent Element
We first need to create a parent element. This will be a list of which our metadata 
inputs can be appended to. 

```{r}
parent_element <- list() 
```

### Publication Date 
The publication date does not have a tab in the excel sheet, but is simply added with 
the `add_pub_date` function. If no date is provided, it will automatically append
the current date. This can be overwritten by providing an input for `date`. 

```{r}
parent_element <- add_pub_date(parent_element = parent_element, date = NULL)
```

### Personnel: Creator, Contact, and Associated Parties. 
The `add_personnel` function allows you to append information on the creator or
the associated parties to your file. The creator would be any person or organization
who is responsible for the creation of the data, while an associated party would 
be any person or organization which is associated to the dataset. The creator will also 
be the contact for this datset. Multiple creators, associated parties, and a mixture
of both can be appended to the EML file by providing all inputs in the "personnel"
tab in the "template.xlsx" excel file. 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
personnel_table <- read_excel("data-raw/template/template.xlsx", sheet = "personnel")
for (i in 1:nrow(personnel_table)) {
  current <- personnel_table[i, ]
  parent_element <- add_personnel(parent_element = parent_element, 
                                  first_name = current$first_name, 
                                  last_name = current$last_name, 
                                  email = current$email, 
                                  role = current$role, 
                                  orcid = current$orcid)
}
```

### Title and Short Name 
Following the personnel element, we can append the title elements. The `add_title` function 
allows you to append the title and short name of the dataset to the file. 
Only one title and short name should be appended. The title should be 
fairly descriptive and between 7 and 20 words long. The short name must be less than
the number of words present in the title and is your opportunity to give viewers a more 
accessible name to the dataset. Again, this information can be added to the "title"
tab in the excel file "template.xlsx".

```{r, }
setwd("~/FlowWest/cvpiaEDIutils")
title <- read_excel("data-raw/template/template.xlsx", sheet = "title")
parent_element <- add_title(parent_element = parent_element, title = title$title, 
                            short_name = title$short_name)
```

### Keyword Set 
The next item we will append is the keyword set. We can use the `add_keyword_set` 
function to do so. The keyword set should include a list of words under a controlled vocabulary, 
that helps better identify your dataset and if it will be applicable to viewers.
A keyword thesaurus is helpful and will provide viewers a better idea of the data,
but is not mandatory. Multiple keyword sets can be provided. This information can 
be added to the "keyword_set" tab in the excel file "template.xlsx".

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
keyword_set <- read_excel("data-raw/template/template.xlsx", sheet = "keyword_set")
thesaurus <- unique(keyword_set$keywordThesaurus)
for (i in 1:length(thesaurus)) {
  keywords <- keyword_set$keyword[keyword_set$keywordThesaurus == thesaurus[i]]
  parent_element <- add_keyword_set(parent_element = parent_element, 
                                    keyword_set = list(keyword = keywords, 
                                                       keywordThesaurus = thesaurus[i]))
}
```

### Abstract
Next, we will use the `add_abstract` function to append the abstract of the dataset to 
your file. The abstract should include basic information on the dataset that gives 
a brief summary to the viewers of what they are observing from the data. This 
information will not be added to the "template.xlsx" excel sheet, but rather the 
"hannon_example_abstract.docx". The text provided can be deleted and replaced with your own. 
From there, the file path of your word document can be added to the excel file 
"template.xlsx". 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
abstract <- read_excel("data-raw/template/template.xlsx", sheet = "abstract")
setwd("~/FlowWest/cvpiaEDIutils")
parent_element <- add_abstract(parent_element = parent_element, 
                               abstract = abstract$abstract)
```

### License and Intellectual Rights
Following the abstract, we will append the license and intellectual rights information. 
The `add_license` function allows you to append the licensing and usage information 
to your file. The `intellectual_rights_description` input calls for the description of
rights, which may be explained in the license you are using. It can also contain 
other requirements for use, other requirements for attribution, etc. Both the CC0 
and CCBY licenses can be chosen as defaults using this function, and the rest of the 
information will be appended for you. Only one license should be appended.
This information can be added to the "license" tab in the excel file "template.xlsx".

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
license <- read_excel("data-raw/template/template.xlsx", sheet = "license")
parent_element <- add_license(parent_element = parent_element, default_license = license$default_license)
```

If a default license is not being used, then the other inputs need to be provided. 
```{r}
# setwd("~/FlowWest/cvpiaEDIutils")
# parent_element <- add_license(parent_element = parent_element, 
#                               license_name = license$license_name,
#                               license_url = license$license_url,
#                               license_identifier = license$license_identifier,
#                               intellectual_rights_description = license$intellectual_rights_description)
```

### Funding
Funding should be appended next. The `add_funding` function allows you to append 
both the description of the funding you have recieved as well as the organization 
you recieved the funding from. Multiple funders can be appended if necessary and 
information should be added to the "funding" tab in the excel file "template.xlsx".

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
funding <- read_excel("data-raw/template/template.xlsx", sheet = "funding")
parent_element <- add_funding(parent_element = parent_element, 
                              funder_name = funding$funder_name,
                              funder_identifier = funding$funder_identifier,
                              award_number = funding$award_number,
                              award_title = funding$award_title,
                              award_url = funding$award_url,
                              funding_description = funding$funding_description)
```

### Maintenance 
The maintenance of a dataset is simply if the data collection is complete or ongoing.
The `add_maintenance` function allows you to append the status of the dataset to
your file. The inputs of `complete` and `ongoing` are the only ones allowed.
If the dataset is still in progress, the frequency of which it is updated must 
be provided as well. This information should be added to the "maintenance" tab 
in the excel file "template.xlsx".

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
maintenance <- read_excel("data-raw/template/template.xlsx", sheet = "maintenance")
parent_element <- add_maintenance(parent_element = parent_element, status = maintenance$status,
                                  update_frequency = maintenance$update_frequency)
```

### Dataset Methods
Overall methods can be appended next. The `add_method` function allows you to
append the different scientific methods used in the collection process of your data.
Methods are appended in a similar fashion as the abstract. A template methods document
is given ("hannon_example_methods.docx). If more methods are needed, you can create 
seperate word documents and provide all of the file paths in the "template.xlsx" 
excel sheet. The methods can be applied here as well as in the data table element,
so make sure you are appending the correct methods to the accurate places. 
Top level methods can describe how a particular csv data was collected, while a data table
method can explain how particular attributes in the data table were collected.

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
method_table <- read_excel("data-raw/template/template.xlsx", sheet = "methods")
for (i in 1:nrow(method_table)) {
  setwd("~/FlowWest/cvpiaEDIutils")
  current <- method_table[i, ]
  new_method <- add_method(parent_element = parent_element, methods_file = current$methods_file,
                           instrumentation = current$instrumentation)
  if (is.null(parent_element$methods)) {
    parent_element$methods <- new_method
  } else {
    parent_element$methods <- list(parent_element$methods, new_method)
  }
}
```

### Coverage: Geographic, Temporal, Taxonomic 
Next, the coverage information is appended. The `add_coverage` function allows
you to append full coverage information to your file. The function itself 
requires inputs for both temporal and geographic coverage. Taxonomic coverage, 
however, is optional and can be appended using the `add_taxonomic_coverage` 
function. Chinook, Delta Smelt, Steelhead, White Sturgeon, and Green Sturgeon are
all default options, with full taxonomic coverage already present for easy appendage
as these are all frequent species being studied under CVPIA. If taxonomic coverage 
is being appended, this information needs to be appended first. It is not appended
to the `parent_element`, however, and is instead appended to a new empty list that
we title `taxonomic_coverage`. The taxonomic coverage information can still be 
added to the "template.xlsx" excel file, even though the information is not 
being appended to the `parent_element`. 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
taxonomic_coverage <- list()
taxonomic_coverage_table <- read_excel("data-raw/template/template.xlsx", sheet = "taxonomic_coverage")
for (i in 1:nrow(taxonomic_coverage_table)) {
  current <- taxonomic_coverage_table[i, ]
  new_taxon <- add_taxonomic_coverage(CVPIA_common_species = current$CVPIA_common_species,
                                     kingdom_value = current$kingdom_value,
                                     phylum_value = current$phylum_value,
                                     class_value = current$class_value,
                                     order_value = current$order_value,
                                     family_value = current$family_value,
                                     genus_value = current$genus_value,
                                     species_value = current$species_value,
                                     common_name = current$common_name,
                                     taxon_id = current$taxon_id)
  if (is.null(taxonomic_coverage)) {
    taxonomic_coverage <- new_taxon
  } else {
    taxonomic_coverage <- append(taxonomic_coverage, new_taxon)
  }
}
```
From here, we then have to append the coverage information to the `parent_element`.
There are three options of doing so.

#### No Taxonomic Coverage 
The first option is that we have no taxonomic coverage present. This can take either
one or multiple coverage inputs and append to the `parent_element`.
```{r}
setwd("~/FlowWest/cvpiaEDIutils")
coverage_table <- read_excel("data-raw/template/template.xlsx", sheet = "coverage")
for (i in 1:nrow(coverage_table)) {
  current <- coverage_table[i, ]
  parent_element <- add_coverage(parent_element = parent_element,
                                 geographic_description = current$geographic_description,
                                 west_bounding_coordinate = current$west_bounding_coordinate,
                                 east_bounding_coordinate = current$east_bounding_coordinate,
                                 north_bounding_coordinate = current$north_bounding_coordinate,
                                 south_bounding_coordinate = current$south_bounding_coordinate,
                                 begin_date = current$begin_date,
                                 end_date = current$end_date)
} 
```
#### Unique Taxonomic Coverage For Multiple Coverages:
If we have multiple coverages, and a unique taxonomic coverage that goes along with 
each specific coverage, we can use this instead: 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
coverage_table <- read_excel("data-raw/template/template.xlsx", sheet = "coverage")
for (i in 1:nrow(coverage_table)) {
  current <- coverage_table[i, ]
  tax <- taxonomic_coverage[i]
  parent_element <- add_coverage(parent_element = parent_element,
                                 geographic_description = current$geographic_description,
                                 west_bounding_coordinate = current$west_bounding_coordinate,
                                 east_bounding_coordinate = current$east_bounding_coordinate,
                                 north_bounding_coordinate = current$north_bounding_coordinate,
                                 south_bounding_coordinate = current$south_bounding_coordinate,
                                 begin_date = current$begin_date,
                                 end_date = current$end_date,
                                 taxonomic_coverage = tax)
} 
```

#### Taxonomic Coverage For Total Coverage 
Finally, if we have multiple taxonomic coverages that need to be appended to one 
overall coverage, we can do the following: 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
coverage_table <- read_excel("data-raw/template/template.xlsx", sheet = "coverage")
for (i in 1:nrow(coverage_table)) {
  current <- coverage_table[i, ]
  parent_element <- add_coverage(parent_element = parent_element,
                                 geographic_description = current$geographic_description,
                                 west_bounding_coordinate = current$west_bounding_coordinate,
                                 east_bounding_coordinate = current$east_bounding_coordinate,
                                 north_bounding_coordinate = current$north_bounding_coordinate,
                                 south_bounding_coordinate = current$south_bounding_coordinate,
                                 begin_date = current$begin_date,
                                 end_date = current$end_date,
                                 taxonomic_coverage = taxonomic_coverage)
} 
```

### Data Table 
Next, we need to create the data table. `physical`, `attribute_list`, and potentially 
`methods` are all lists which must be appended to the data table list, so we 
have to create these first. 

#### Physical 
The physical element can be created first. This represents information on the 
actual physical information of the data. The file path of the data can be added 
to the "template.xlsx" excel file and all parameters will be appended, gathering
the csv's unique size and authentication as well as providing default values of a csv. 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
physical_list <- list()
physical_table <- read_excel("data-raw/template/template.xlsx", sheet = "physical")
for (i in 1:nrow(physical_table)) {
  current <- physical_table[i,]
  new_physical <- add_physical(file_path = current$file_path, 
                               data_url = current$data_url)
  if (is.null(physical_list$physical)) {
    physical_list$physical <- new_physical
  } else {
    physical_list$physical <- list(physical_list$physical, new_physical)
  }
}
```
If you wish to provide alternative information than the defaults of a csv, you must do
so manually in Rstudio using the `add_physical` function fully, although it is not 
recommended that you do so. 

```{r}
# physical_list <- list()
# physical_list <- add_physical(file_path = file_path, 
#                                number_of_headers = number_of_headers,
#                                record_delimiter = record_delimiter,
#                                attribute_orientation = attribute_orientation,
#                                field_delimiter = field_delimiter,
#                                data_url = data_url)
```

#### Attribute List 
Next, we will need to append an attribute list to the data table. We will be able 
to use the `add_attribute` function to  do so. Please make sure you review what 
type of attribute you are providing and what inputs are necessary. These values then 
can be inputted into the "attribute" tab in the "template.xlsx" excel file. If you
are using a "nominal" or "ordinal" attribute which is "enumerated", in other words
it has a code definition, please also use the tab "code_definitions". Provide each
unique code and definition, with its `attribute_name` alligning to that which is in the
"attirbute" tab. An example is present curently to help better showcase this. 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
attribute_list <- list()
codes <- read_excel("data-raw/template/template.xlsx", sheet = "code_definitions")
attribute_names <- unique(codes$attribute_name)
attribute_table <- read_excel("data-raw/template/template.xlsx", sheet = "attribute")
for (i in 1:nrow(attribute_table)) {
  current <- attribute_table[i, ]
  if (current$domain %in% "enumerated") { 
    definition <- list()
    current_codes <- codes[codes$attribute_name == current$attribute_name, ]
    for (j in 1:nrow(current_codes)) {
      codeDefinition <- list(code = current_codes$code[j], definition = current_codes$definitions[j])
      if (is.null(definition$codeDefinition)) {
        definition$codeDefinition <- codeDefinition
      } else {
        definition$codeDefinition <- list(definition$codeDefinition, codeDefinition)
      }
    }
  } else {
    definition = current$definition
  }
  new_attribute <- add_attribute(attribute_name = current$attribute_name,
                                 attribute_definition = current$attribute_definition,
                                 storage_type = current$storage_type,
                                 measurement_scale = current$measurement_scale,
                                 domain = current$domain,
                                 definition = definition,
                                 text_pattern = current$text_pattern,
                                 type = current$type,
                                 units = current$units,
                                 unit_precision = current$unit_precision,
                                 number_type = current$number_type,
                                 date_time_format = current$date_time_format,
                                 date_time_precision = current$date_time_precision,
                                 minimum = current$minimum,
                                 maximum = current$maximum,
                                 attribute_label = current$attribute_label)
  if (is.null(attribute_list$attribute)) {
    attribute_list$attribute <- new_attribute
  } else {
    attribute_list$attribute <- list(attribute_list$attribute, new_attribute)
  }
}
```

### Methods
Finally, if there are any methods which you would like to include at the data table
level, this is where you would do this. It is optional, but as stated before, can 
be included in both places if necessary. Like the previous methods, a document 
needs to be created which holds all of your methods, and the file path can be added 
to the "data_table_methods" tab in the "template.xlsx" excel sheet. 

```{r}
setwd("~/FlowWest/cvpiaEDIutils")
method_list <- list()
data_table_methods <- read_excel("data-raw/template/template.xlsx", sheet = "data_table_methods")
for (i in 1:nrow(data_table_methods)) {
  setwd("~/FlowWest/cvpiaEDIutils")
  current <- data_table_methods[i, ]
  new_method <- add_method(parent_element = method_list, methods_file = current$methods_file,
                            instrumentation = current$instrumentation)
  if (is.null(method_list$methods)) {
    method_list$methods <- new_method
  } else {
    method_list$methods <- list(method_list$methods, new_method)
  }
}
```

### Putting the Data Table Together
Now that we have the `attribute_list`, `physical` information, and `methods`, we 
can compose the data table, which is the last element we need to create our 
working EML file. The remaining inputs for the data table can be put in the "data_table"
tab in the "template.xlsx" excel file. Multiple data tables can be appended to one dataset,
although it is more common to only have one present. 
```{r}
setwd("~/FlowWest/cvpiaEDIutils")
data_table <- read_excel("data-raw/template/template.xlsx", sheet = "data_table")
for (i in 1:nrow(data_table)) {
  current <- data_table[i, ]
  phy <- physical_list[i]
  att <- attribute_list[i]
  method <- method_list[i]
  parent_element <- add_data_table(parent_element = parent_element,
                                   entity_name = current$entity_name,
                                   entity_description = current$entity_description,
                                   physical = phy,
                                   attribute_list = att,
                                   methods = method,
                                   number_of_records = current$number_of_records,
                                   alternate_identifier = current$alternate_identifier)
}
```
If methods are not present at this level, make sure to take it out. 
```{r}
setwd("~/FlowWest/cvpiaEDIutils")
data_table <- read_excel("data-raw/template/template.xlsx", sheet = "data_table")
for (i in 1:nrow(data_table)) {
  current <- data_table[i, ]
  phy <- physical_list[i]
  att <- attribute_list[i]
  parent_element <- add_data_table(parent_element = parent_element,
                                   entity_name = current$entity_name,
                                   entity_description = current$entity_description,
                                   physical = phy,
                                   attribute_list = att,
                                   number_of_records = current$number_of_records,
                                   alternate_identifier = current$alternate_identifier)
}
```

## Putting It All Together
Now that we have all of the items appended to the `parent_element`, we must appended it 
to the "dataset" and also give the file a unique ID number. 
```{r}
eml <- list(packageId = "uuid::UUIDgenerate()",
            system = "uuid",
            dataset = parent_element)
```
We now have everything we need present in the EMML document. To get it into the 
correct format we can use the EML library's `write_eml` function. 
```{r}
EML::write_eml(eml, "eml.xml")
```
