---
title: "Dataset Element"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataset-Element}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
The following libraries are needed to create a working EML document. 
```{r, message = FALSE}
library(EDIutils)
library(tidyverse)
library(readxl)
library(EML)
```

# Dataset Element 
This document explains how to add multiple datasets to one EML document and how to 
add non tabular data to an EML document. 

## Loading in Data 
```{r, warning=F, include=FALSE}
excel_path <- system.file("extdata", "Hannon-Example", "example-metadata.xlsx", 
                          package = "EDIutils", mustWork = TRUE)

sheets <- readxl::excel_sheets(excel_path)
metadata <- lapply(sheets, function(x) readxl::read_excel(excel_path, sheet = x))
names(metadata) <- sheets

abstract_docx <- system.file("extdata", "Hannon-Example", "hannon_example_abstract.docx", 
                             package = "EDIutils", mustWork = TRUE)

methods_docx <- system.file("extdata", "Hannon-Example", "hannon_example_methods.docx", 
                            package = "EDIutils", mustWork = TRUE)

dataset_file = system.file("extdata", "Hannon-Example", "hannon_example_physical_data.csv",
                            package = "EDIutils", mustWork = TRUE)

dataset_file_name <- "hannon_example_physical_data.csv"

edi_number = "edi.687.1"
```

```{r, eval = FALSE}
excel_path <- "data-raw/Hannon-Example/example-metadata.xlsx"

sheets <- readxl::excel_sheets(excel_path)
metadata <- lapply(sheets, function(x) readxl::read_excel(excel_path, sheet = x))
names(metadata) <- sheets

abstract_docx <- "data-raw/Hannon-Example/hannon_example_abstract.docx"

methods_docx <- "data-raw/Hannon-Example/hannon_example_methods.docx"

dataset_file <- "data-raw/Hannon-Example/hannon_example_physical_data.csv"

dataset_file_name <- "hannon_example_physical_data.csv"

edi_number <- "edi.678.1"
```

## Multiple datsets for one data package
In this example we included two datasets by creating a tibble where each row gives information on one dataset. 

The tibble has 3-5 columns:

* A `datatable` element which is the file path to the datatable
* A `datatable_name`
* An `attribute_info` the file path for the attribute and code tables for the dataset (should be a metadata.xlsx)
* (Optional) `dataset_methods` section that reads file path to dataset specific methods
* (Optional) `additional_information` If additional information is pertinent to the entry please add it as a string description. 

In this example there are no dataset specific methods or additional information so those columns are not included. 

To add more than two datasets please follow the same structure adding additional an additional row for each dataset. 

```{r}
# TODO add real multipule datast example 
dataset_files <- dplyr::tibble(datatable = c(system.file("extdata", "Hannon-Example",
                                                                    "hannon_example_physical_data.csv",
                                                        package = "EDIutils", mustWork = TRUE),
                                             system.file("extdata", "Hannon-Example",
                                                                    "hannon_example_physical_data.csv",
                                                        package = "EDIutils", mustWork = TRUE)),
                               datatable_name = c("hannon_example_snorkel_data.csv", 
                                                  "hannon_example_physical_data.csv"),
                               attribute_info = c(system.file("extdata", "Hannon-Example", "example-metadata.xlsx",
                                                       package = "EDIutils", mustWork = TRUE),
                                                  system.file("extdata", "Hannon-Example", "example-metadata.xlsx",
                                                       package = "EDIutils", mustWork = TRUE)))

```

### Data Tables 
We need to add each data table. The data table element includes the datatable name,  a `physical` section, and the `attribute_list`. These sections are all lists which must be created first then added to a named `dataTable` list.

The code below create a `data_tables` list where we append each data table. We have the `code_helper` and the `attributes_and_codes` that help create the  `attribute_list` using the `add_attribute` function. We create the `physical` 
element using the `add_physical` function. 

Please make sure you review what type of attribute you are providing and what inputs are necessary. 
These values can then be inputted into the "attribute" tab in the "example-metadata.xlsx" excel file. If you
are using a "nominal" or "ordinal" attribute which is "enumerated", (it has a specific 
code definition), please also use the tab "code_definitions". Provide each
unique code and definition, with its `attribute_name` aligning to that which is in the
"attribute" tab. An example is present currently to help better showcase this. Every single 
column in the dataTable must have a described attribute to match EDI congruence checker. 

```{r}
# TODO work in dataset methods 
adds_datatable <- function(datatable, datatable_name, attribute_info, dataset_methods = NULL, additional_info = NULL){
  attribute_table <- readxl::read_xlsx(attribute_info, sheet = "attribute")
  codes <- readxl::read_xlsx(attribute_info, sheet = "code_definitions")
  attribute_list <- list()
  attribute_names <- unique(codes$attribute_name)
  
  # Code helper function 
  code_helper <- function(code, definitions) {
  codeDefinition <- list(code = code, definition = definitions)
  }
  # Attribute helper function to input into pmap
  attributes_and_codes <- function(attribute_name, attribute_definition, storage_type, 
                                   measurement_scale, domain, type, units, unit_precision, 
                                   number_type, date_time_format, date_time_precision, minimum, maximum, 
                                   attribute_label){
    if (domain %in% "enumerated") { 
      definition <- list()
      current_codes <- codes[codes$attribute_name == attribute_name, ]
      definition$codeDefinition <- purrr::pmap(current_codes %>% select(-attribute_name), code_helper) 
    } else {
      definition = attribute_definition
    }
    new_attribute <- add_attribute(attribute_name = attribute_name, attribute_definition = attribute_definition,
                                   storage_type = storage_type, measurement_scale = measurement_scale, 
                                   domain = domain, definition = definition, type = type, units = units, 
                                   unit_precision = unit_precision, number_type = number_type, 
                                   date_time_format = date_time_format, date_time_precision = date_time_precision, 
                                   minimum = minimum, maximum = maximum, attribute_label = attribute_label)
  }
  attribute_list$attribute <- purrr::pmap(attribute_table, attributes_and_codes)
  
  physical <- add_physical(file_path = datatable)
  dataTable <- list(entityName = datatable_name,
                    entityDescription = metadata$dataset$name,
                    physical = physical,
                    attributeList = attribute_list)
}
data_tables <- purrr::pmap(dataset_files, adds_datatable) 
```
This `data_table` list can now be added in as `dataTable = data_tables` in the dataset list. 

## Diffrent types of datasets 
##### Physical 
The physical element can be created first using the `add_physical` function. 
This will append information on the actual physical information of the data. 
The file path of the data can be added to the "example-metadata.xlsx" excel file 
and all parameters will be appended, gathering the csv's unique size and 
authentication, as well as providing the default physical values of a csv. 

```{r}
physical <- add_physical(file_path = dataset_file)
```

If you wish to provide alternative information than the defaults of a csv, you must do
so manually in Rstudio using the `add_physical` function fully. It is not recommended 
that you do so, however, as the default values present represent the best practices 
in most cases. 

### Vector Data 
If you have vector data please append the data elements in a `spatialVector` named section of the dataset. In addition to the components that are required for tabular data you must include a geometry for your dataset. This should be added in the "dataset tab" of the "example-metadata.xlsx" excel file.
```{r, eval = FALSE}
#TODO add real example of spatialVEctor
data_tables <- list(spatialVector = list(entityName = dataset_file,
                                         entityDescription = metadata$dataset$name,
                                         physical = physical,
                                         attributeList = attribute_list,
                                         geometry = metadata$dataset$geometry))
```

### Raster Data 
If you have raster data please append the data elements in a `spatialRaster` named section of the dataset. In addition to the components that are required for tabular data you must include a geometry for your dataset. This should be added in the "dataset tab" of the "example-metadata.xlsx" excel file.
```{r, eval = FALSE}
# TODO add real example for spatialVector 
data_tables <- list(spatialRaster = list(add_raster()))
```

#### Checks for valid EML
```{r, include = FALSE}
dataset <- list(title = "Testing for valid EML on multiple datasets",
                creator = list(individualName = list(givenName = "John M.", 
                                                     surName = "Hannon"), 
                               electronicMailAddress = "jhannon@usbr.gov", 
                               organizationName = "US Bureau of Reclamation"),
                contact =list(individualName = list(givenName = "John M.", 
                                                     surName = "Hannon"), 
                               electronicMailAddress = "jhannon@usbr.gov", 
                               organizationName = "US Bureau of Reclamation"),
                dataTable = data_tables) 
```

```{r, include = FALSE}
eml <- list(packageId = edi_number,
            system = "EDI",
            dataset = dataset)
```
The final step is to convert our `eml` list into the correct format. To do so, we can use the EML library's `write_eml` function. 
```{r, eval = FALSE, include = FALSE}
file_name <- paste(edi_number, "xml", sep = ".")
EML::write_eml(eml, file_name)
eml_validate(file_name)

```