---
title: "Details on Inital Metadata Components"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inital-Metadata-Information}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
The following libraries are needed to create a working EML document. 
```{r, message = FALSE}
library(EDIutils)
library(tidyverse)
library(readxl)
library(EML)
```

```{r, include = FALSE}
excel_path <- system.file("extdata", "Hannon-Example", "example-metadata.xlsx", 
                          package = "EDIutils", mustWork = TRUE)

sheets <- readxl::excel_sheets(excel_path)
metadata <- lapply(sheets, function(x) readxl::read_excel(excel_path, sheet = x))
names(metadata) <- sheets

abstract_docx <- system.file("extdata", "Hannon-Example", "hannon_example_abstract.docx", 
                             package = "EDIutils", mustWork = TRUE)

methods_docx <- system.file("extdata", "Hannon-Example", "hannon_example_methods.docx", 
                            package = "EDIutils", mustWork = TRUE)

dataset_file <- system.file("extdata", "Hannon-Example", "hannon_example_physical_data.csv",
                               package = "EDIutils", mustWork = TRUE)

dataset_file_name <- "hannon_example_physical_data.csv"

edi_number <- "edi.678.1"
```
# Custom inputs to the initial metadata information 

This document gives more in depth examples on how to use the `add_access`, `add_pub_date`,
`add_keywords`, and `add_license` elements. This document will not explain how to create a complete valid EML document to upload to the EDI repository. Please use this article just for sections of the EML document and refer to [the EML template helper](https://cvpia-osc.github.io/EDIutils//articles/creating-EML.html) to generate the entire document.

## Access Permissions 
### Default Permissions
The `add_access` adds an access section at the beginning of our EML document. The 
`add_access` default is public principal with a read permission. 

```{r}
access <- add_access()
```

### Custom Permissions 
If you would like to change the principle and permissions for access please specify a 
valid principle and permission. 
The valid options for principal are:

  * 'public' - special principal indicate that any user or group has a particular access permission   
  * other individual users or groups can be specified if they are defined in the PASTA EDI authentication 
   system. Read the [PASTA documentation](https://pastaplus-core.readthedocs.io/_/downloads/en/latest/pdf/) for more information 
   
The valid options for permission are:  

  * 'read' - allow viewing of the resource, 
  * 'write' - allow modification of the resource (except for access rules),   
  * 'changePermission' - modifications including access rules, and  
  * 'all' - all of the above.

The example below shows the `add_access` function taking in a public principle with a permission of all. 
```{r}
access <- add_access(allow_principal = "public", allow_permission = "all")
```

## Publication Date 
### Default Publication Date
The publication date does not have a tab in the excel sheet, but is simply added with 
the `add_pub_date` function. If no date is provided, it will automatically append
the current date. This can be overwritten by providing an input for `date`. 

```{r}
publication_date <- add_pub_date(list())
```

### Custom Publication Date 
If you wish to provide an input for date please do so using a date in international date 
format (ISO). A date in ISO format lists year, month and day as follows: "YYYY-MM-DD". The example 
below uses `add_pub_date` to add a custom date. 
```{r}
publication_date <- add_pub_date(list(), date = "2020-08-19")
```

## Keyword Set 
### Keywords that do not come from a controlled vocabulary 
The next item we will append is the keyword set. We can use the `add_keyword_set` 
function to do so. The keyword set should include a list of words that help 
identify your project or connect it with other similar projects. This information can 
be added to the "keyword_set" tab in the excel file "example-metadata.xlsx".

```{r}
keyword_set <- metadata$keyword_set
keywords <- add_keyword_set(list(), keyword_set[,1])
```

### Keywords from a controlled vocabulary
If you add keywords
from a controlled vocabulary you must state what keyword thesaurus they belong to. 
Multiple keyword sets can be provided. This information can 
be added to the "keyword_set" tab in the excel file "example-metadata.xlsx".

```{r, eval = FALSE}
unique_thesaurus <- unique(metadata$keyword_set$keywordThesaurus)
keywords <- list()
add_keywords = function(unique_thesaurus){
  keywords <- add_keyword_set(keywords, 
                              keyword_set = list(keyword = metadata$keyword_set$keyword[keyword_set$keywordThesaurus == unique_thesaurus],
                                                 keywordThesaurus = unique_thesaurus))
}
all_keywords <- purrr::map(unique_thesaurus, add_keywords) %>% flatten()
```

## License and Intellectual Rights
### Default Licenses 
The `add_license` function allows you to append the licensing and usage information 
to your file.  Both the CC0 and CCBY licenses can be chosen as defaults using 
this function, and the rest of the information will be appended for you. This information 
can be added to the "license" tab in the excel file "example-metadata.xlsx".

```{r}
license <- add_license(list(), 
                       default_license = metadata$license$default_license)

```

### Custom Licenses 
If a default license is not being used, then the other inputs need to be provided in 
the "license" tab in the excel file "example-metadata.xlsx". 
```{r, eval = FALSE}
license <- add_license(list(),
                       license_name = metadata$license$license_name,
                       license_url = metadata$license$license_url,
                       license_identifier = metadata$license$license_identifier,
                       intellectual_rights_description = metadata$license$intellectual_rights_description)

```






