---
title: "Update Data Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Update Data Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(EMLaide)
```

## Updates to the EDI portal

This article describes how to update an EDI package using EMLaide. See the official EDI guidelines for when updating a package is appropriate at the link [here](https://edirepository.org/resources/updating-a-data-package).

EDI updates are very similar to data package uploads.  

## Generate EML Document

To update an existing EDI package, you will have to first create a new EML with the updated tables and metadata. Please reference the article [Create EML from Template](https://cvpia-osc.github.io/EMLaide/articles/create-EML-from-template.html) to see specifications for creating an EML document using EMLaide.

### Updating the EDI identifier

When creating the EML document for an update (as opposed to a new EDI package), you must update your EDI identifier to reflect the version of the package:

```{r}
old_edi_number <- "edi.750.1" 
updated_edi_number <- "edi.750.2"
```

This will be reflected when you write the updated XML:

```{r, eval = FALSE}
eml_updated <- list(packageId = updated_edi_number,
                   system = "EDI",
                   access = add_access(),
                   dataset = dataset,
                   additionalMetadata = list(metadata = list(unitList = unitList)))

# write and validate updated EML document
EML::write_eml(eml_updated, paste0(updated_edi_number, ".xml"))
EML::eml_validate(paste0(updated_edi_number, ".xml"))
```

## Permissions for updates

You must have the credentials (EDI username, EDI password) for the account that uploaded the original EDI package in order to update it.

## Update the package

### Update manually via portal

One option is to update the package manually on the [EDI portal](https://portal.edirepository.org/nis/login.jsp). The steps to updating your package on the portal are as follows: 

1) Login using the credentials associated with the original EDI package
2) Navigate to *Tools* --> *Evaluate/Upload Data Package*
3) Upload your updated data package according to the instructions
4) Ensure that the edi number reflects the updated versioning
5) If applicable, under **Data Upload Options**, select *Allow PASTA+ to skip the upload of a data entity if it has a matching copy*.


## Update using API 

Assuming you have written your updated EML document with the identifier `updated_edi_number`, you can evalutae and update your package from RStudio using an API.

To evaluate your updated edi package, use the function `EMLaide::evaluate_edi_package()`. To update it, use the function `EMLaide::update_edi_package()`.

The following code evaluates a data package with identifier `updated_edi_number` and updates it by passing the updated .xml file and the past version's identifier `old_edi_number`:

```{r, eval = FALSE}
# test that the package passes EDIs evaluation criteria 
evaluate_edi_package(user_id = "your user id", 
                     password = "your user password",
                     eml_file_path = paste0(updated_edi_number, ".xml"), 
                     environment = "production")

# update edi package 
update_edi_package(user_id = "your user id", 
                   password = "your user password",
                   eml_file_path = paste0(updated_edi_number, ".xml"),
                   package_size = "medium", # default
                   existing_package_identifier = old_edi_number, 
                   environment = "production")
```

